Bootstrapping Multi-atlas Hippocampal Segmentation with \mb

### 2013-08-26 paper.Rnw
Jen's modifications

The hippocampus is a neuroanatomical structure situated in the medial temporal
lobe of the brain, and has long been associated with learning and memory
\citep{DenHeijer2012,Scoville2000}. In addition to its known functional roles,
the hippocampus is of interest to neuroscientists because it is implicated in
several forms of brain disfunction such as Alzheimer's disease
\citep{Sabuncu2011} and schizophrenia \citep{Narr2004,Karnik-Henry2012}. In
neuroimaging experiments, magnetic resonance images (MRI) are often used for the
identification of the hippocampus. As such, accurate segmentation of the
hippocampus and its subregions in MRI is a necessary first step to better
understand the unique neuroanatomy of subjects. Typically, the gold standard for
neuroanatomical segmentation is manual delineation by an expert human rater.
However with the availability of increasingly large MRI datasets, the time and
expertise required for manual segmentation becomes prohibitive
\citep{Mazziotta1995,Mazziotta2001,Mazziotta,Pausova2007}. Further, there is
little agreement between researchers regarding how exactly the hippocampus
should be identified in MRI images \citep{Geuze2004} and this has led to efforts
to create an unified hippocampal segmentation protocol \citep{Jack2011}.

Automated segmentation techniques that are reliable, objective, and reproducible
are a necessary alternative to manual segmentation. In the case of classical
model-based segmentation methods \citep{Haller1997,Csernansky1998}, an MRI atlas
that was previously manually labelled by an expert rater is matched to target
images using nonlinear registration methods. The resulting nonlinear
transformation is applied to the manual labels (i.e. {\em label propagation}) to
move them into the target image space. While this methodology has been used
successfully in several contexts
\citep{Chakravarty2008,Chakravarty2009,Collins1995,Haller1997}, it is limited in
accuracy due to inaccuracies in the estimated nonlinear transformation itself,
partial volume effects in label resampling, and irreconcilable differences
between the neuroanatomy represented within the atlas and target images.

@@

To the best of our knowledge there are two other segmentation methods that
attempt to define the hippocampus and its subfields using standard T1-weighted
data. The first is included with the FreeSurfer package \citep{VanLeemput2009}.
This work is limited as it omits the tail of the hippocampus and the
segmentation protocol has yet to be fully validated. Nonetheless, it
demonstrates that the applicability of hippocampal subfield segmentation using
data from 10 subjects. In the second method, \citet{Yushkevich2009} acquired
and labelled hippocampal subfields on high-resolution MRI data from post-mortem
medial temporal lobe samples. Using nonlinear registration guided by manually
derived hippocampus masks and specific landmarks, they demonstrate accurate
parcellation of hippocampal subfields in unlabelled MRI volumes.

>>>>

The hippocampus is a brain structure situated in the medial temporal
lobe, and has long been associated with learning and memory
\citep{DenHeijer2012,Scoville2000}. The hippocampus is of interest to
neuroscientists because it is implicated in many forms of brain disfunction,
including Alzheimer's disease \citep{Sabuncu2011} and schizophrenia
\citep{Narr2004,Karnik-Henry2012}. In neuroimaging studies, structural magnetic
resonance images (MRI) are often used for the volumetric assessment of the
hippocampus.  As such, accurate segmentation of the hippocampus and its
subregions in MRI is a necessary first step to better understand the
inter-individual variability of subject neuroanatomy. Typically, the gold
standard for neuroanatomical segmentation is manual delineation by an expert
human rater.  However, with the availability of increasingly large MRI datasets,
the time and expertise required for manual segmentation becomes prohibitive
\citep{Mazziotta1995,Mazziotta2001,Mazziotta,Pausova2007}. Further, there is
significant variation between segmentation protocols regarding specific
anatomical boundaries of the hippocampus \citep{Geuze2004} and this has led to
efforts to create an unified hippocampal segmentation protocol \citep{Jack2011}.

Automated segmentation techniques that are reliable, objective, and reproducible
can be considered complementary to manual segmentation. In the case of classical
model-based segmentation methods \citep{Haller1997,Csernansky1998}, an MRI atlas
that was previously manually labelled by an expert rater is matched to target
images using nonlinear registration methods. The resulting nonlinear
transformation is applied to the manual labels (i.e. {\em label propagation}) to
move them into the target image space. While this methodology has been used
successfully in several contexts
\citep{Chakravarty2008,Chakravarty2009,Collins1995,Haller1997}, it is limited in
accuracy due to inaccuracies in the estimated nonlinear transformation itself,
partial volume effects in label resampling, and irreconcilable differences
between the neuroanatomy represented within the atlas and target images.

@@

To the best of our knowledge there are only two other segmentation methods that
attempt to define the hippocampus and its subfields using standard T1-weighted
data. The first is included with the FreeSurfer package \citep{VanLeemput2009}.
This work is limited as it omits the tail of the hippocampus and the
segmentation protocol has yet to be fully validated. Nonetheless, it
demonstrates that the applicability of hippocampal subfield segmentation using
data from 10 subjects. In the second method, \citet{Yushkevich2009} hippocampal
subfields were acquired and labelled on high-resolution MRI data from post-mortem
medial temporal lobe samples. Using nonlinear registration guided by manually
derived hippocampus masks and specific landmarks, the authors demonstrated
accurate parcellation of hippocampal subfields in unlabelled MRI volumes.
